Diagnostic Potential & the Interactive Dynamics of the Colorectal Cancer Virome

### 2018-05-27 doc/manuscript.Rmd
Final doc edits

Human viruses (those that infect human cells) have been associated with many cancers, largely due to their mutagenic and functionally manipulative abilities. Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses. We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions but an unknown association with the virome. We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition. Through random forest modeling we identified differences in the healthy and colorectal cancer virome. The cancer-associated virome consisted primarily of temperate bacteriophages that were also predicted to be bacteria-virus community network hubs. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.

@@

# Introduction
The human gut virome is the community of all viruses found in the gut, including bacteriophages (viruses that only infect bacteria) and eukaryotic viruses (viruses that only infect eukaryotic cells), and human-specific viruses (viruses that only infect human cells). Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages, another major component of the virome, are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and members of those bacterial communities have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer as well. The gut virome therefore has a potential to be associated with, and potentially impact, human cancer. Altered human virome composition and diversity have already been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The strong association of bacterial communities with colorectal cancer, the previous identification of human-specific viruses that cause cancer, and the precedent for the virome to impact other human diseases suggest that colorectal cancer may be associated with altered virus communities.

@@

Here we begin to address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those differences might impact cancer progression and severity. We also aimed to evaluate the virome's potential for use as a diagnostic biomarker. The implications of this study are threefold. *First*, this work supports a biological role for the virome in colorectal cancer development and suggests that more than the bacterial members of the associated microbial communities are involved in the process. *Second*, we present a supplementary, or even alternative, virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.

@@

## Unaltered Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e., "dysbiotic"). Therefore, we first evaluated the influence of colorectal cancer on virome OVU diversity. We evaluated differences in communities between disease states using the Shannon diversity, richness, and Bray-Curtis metrics. We observed no significant alterations in either Shannon diversity or richness in the diseased states as compared to the healthy state **(Figure \ref{betaogu} C-D)**. There was no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**). Notably, there was a significant difference between the few blank controls that remained after rarefying the data and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})**, further supporting the quality of the sample set. In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**. This is consistent with what has been observed when the same metrics were applied to 16S rRNA gene sequences and metagenomic samples [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these communities.

## Virome Composition in Colorectal Cancer
As opposed to the diversity metrics discussed above, OTU-based relative abundance profiles generated from 16S rRNA gene sequences are effective for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja]. The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer. We built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared their performance to models built using bacterial community signatures.

@@

We compared viral metagenomic methods to bacterial metagenomic methods by building a viral model and a model built using OGU relative abundance profiles from bacterial metagenomic shotgun sequencing data. This bacterial model performed worse than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = asigfig)`) **(Figure \ref{predmodel} A - B)**. To determine the cause of the discrepency between the two bacterial sequencing methods, we attempted to compare the approaches at a common sequencing depth. This revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**. Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% across the samples **(Figure \ref{16scompare} B)**. Although the features in the viral model also were of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.

The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature also was driven by a few OVUs, suggesting a role for these viruses in tumorigenesis **(Figure \ref{predmodel} D)**. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and "unclassified" phage taxa. Many of the important viruses were unidentifiable (denoted "unknown"). This is common in viromes across habitats; studies have reported as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

## Phage Influence Between CRC Stages
Because previous work has identified shifts in which bacteria were most important at different stages of colorectal cancer [@Zackular:2014fba; @Zackular:2016en; @Flynn:2016iu], we explored whether shifts in the relative influence of phages could be detected between healthy, adenomatous, and cancerous colons. We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions. While bacterial OTU models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important between the healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.

@@

## Phage Dominance in CRC Virome
Differences in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes. It is important to note that this could have been influenced by our methodological biases against enveloped viruses (more common of eukaryotic viruses than bacteriophage), due to chloroform and DNase treatment for purification.

@@

Given these findings, we hypothesized that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria. In other words, we hypothesized that the influential bacteriophages were community hubs (i.e. central members) within the bacteria and phage interactive network. We investigated the potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community **(Figure \ref{network} A)** [@Hannigan:2017cj]. The predicted interactions were then used to identify phage community hubs. We calculated the alpha centrality (i.e. measure of importance in the ecological network) of each phage OVU's connection to the rest of the network. The phages with high centrality values were defined as community hubs. Next, the centrality of each OVU was compared to its importance in the colorectal cancer classification model. Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer also were more likely to be community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities and had broad host ranges.

@@

Using our findings here, we have begun to delineate the role the colonic virome plays in colorectal cancer in the form of a working hypothesis that will guide our future studies **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states. By implementing a machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer. The colorectal cancer virome was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].

In addition to modeling the potential connections between virus communities, bacterial communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur. This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu]. We hypothesize that the process begins with broadly infectious phages in the colon lysing and thereby disrupting the existing bacterial communities. This shift opens novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize. Once the initial influential founder bacteria establish themselves in the epithelium, secondary opportunistic bacteria are able to adhere to the founders, colonize, and establish a biofilm. Phages may play a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process important for effective biofilm growth [@Rossmann:2015cj]. The oncogenic bacteria may then be able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response. As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria whose relative abundances were most influential. As the bacteria enter their oncogenic synergy with the epithelium, we conjecture that the phages continue mediating biofilm dispersal. This process would thereby support the colonized oncogenic bacteria by lysing competing cells and releasing nutrients to other bacteria in the form of cellular lysates. In addition to highlighting the likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities this model will guide future research investigations of the role the virome plays colorectal cancer.

At this time, our working hypothesis represents a conceptualization of areas for future work, which will be required to characterize the colorectal cancer microbiome at the functional, mechanistic level. There are many different ways in which this system can be operating, and our working hypothesis is one. For example, it is possible that the bacterial communities cause a change in the virome instead of the virome altering the bacterial communities. Reviewing every possible mechanism and characteristic of the system is beyond the scope of this paper, but the results we present provide a step toward the answer. Important future directions will include larger cohort human studies, further *in vitro* and *in vivo* mechanistic experimentation, and even attempts at community studies using absolute abundance values instead of relative abundance, which would allow for more accurate community dynamic modeling.

@@

https://github.com/SchlossLab/Hannigan_CRCVirome_mBio_2017

@@

## Ecological Network Analysis & Correlations
The ecological network of the bacterial and phage operational genomic units was constructed and analyzed as previously described [@Hannigan:2017cj]. Briefly, a random forest model was used to predict interactions between bacterial and phage genomic units, and those interactions were recorded in a graph database using *neo4j* graph databasing software (v2.3.1). The degree of phage centrality was quantified using the alpha centrality metric in the igraph CRAN package. A Spearman correlation was performed between model importance and phage centrality scores.

@@

# Acknowledgments
The authors thank the Schloss lab members for their underlying contributions, and the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.

@@

![*Basic Quality Control Metrics. A) VLP genomic DNA yield from all sequenced samples. Each bar represents a sample which is grouped and colored by its associated disease group. B) Sequence yield following quality control including quality score filtering and human decontamination.*\label{qualcontrol}]()

@@

![*Beta-diversity comparing disease states and the study negative controls. Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test. Comparisons in which the intervals cross the zero mean difference line (dashed line) were not significantly different.*\label{betaogunegative}]()

>>>>

Human viruses (those that infect human cells) have been associated with many cancers, largely due to their mutagenic and functionally manipulative abilities. Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses. We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions but with an unknown association with the gut virome. We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition. Through random forest modeling we identified differences in the healthy and colorectal cancer virome. The cancer-associated virome consisted primarily of temperate bacteriophages that were also predicted to be bacteria-virus community network hubs. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and potentially impact cancer progression by altering the bacterial host communities.

@@

# Introduction
The human gut virome is the community of all viruses found in the gut, including bacteriophages (viruses that only infect bacteria), eukaryotic viruses (viruses that only infect eukaryotic cells), and human-specific viruses (viruses that only infect human cells). Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and members of those bacterial communities have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer as well. The gut virome therefore has a potential to be associated with, and potentially impact, human cancer. Altered human virome composition and diversity have already been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The strong association of bacterial communities with colorectal cancer, the previous identification of human-specific viruses that cause cancer, and the precedent for the virome to impact other human diseases suggest that colorectal cancer may be associated with altered virus communities.

@@

Here we begin to address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those differences might impact cancer progression and severity. We also aimed to evaluate the virome's potential for use as a diagnostic biomarker. The implications of this study are threefold. *First*, this work supports a biological role for the virome in colorectal cancer development and suggests that more than the bacterial members of the associated microbial communities are involved in the process. *Second*, we present a supplementary virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.

@@

## Unaltered Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiotic"). Therefore, we first evaluated the influence of colorectal cancer on virome OVU diversity. We evaluated differences in communities between disease states using the Shannon diversity, richness, and Bray-Curtis metrics. We observed no significant alterations in either Shannon diversity or richness in the diseased states as compared to the healthy state **(Figure \ref{betaogu} C-D)**. There was no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**). Notably, there was a significant difference between the few blank controls that remained after rarefying the data and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})**, further supporting the quality of the sample set. In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**. This is consistent with what has been observed when the same metrics were applied to 16S rRNA gene sequences and metagenomic samples [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these community structures.

## Virome Composition in Colorectal Cancer
As opposed to the diversity metrics discussed above, OTU-based relative abundance profiles generated from 16S rRNA gene sequences are effective for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja]. By using classification models instead of attempting to identify single differentially abundant OTUs, these and other studies have been successful in capturing complex community relationships in which differences in taxonomic relative abundance are considered in the context of other taxa. The exceptional performance of bacteria in these classification models supports a role for bacterial functionality in colorectal cancer. We built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared their performance to models built using bacterial community signatures.

@@

We compared viral metagenomic methods to bacterial metagenomic methods by building a viral model and a model built using OGU relative abundance profiles from bacterial metagenomic shotgun sequencing data. This bacterial model performed worse than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = asigfig)`) **(Figure \ref{predmodel} A - B)**. To determine the cause of the discrepancy between the two bacterial sequencing methods, we attempted to compare the approaches at a common sequencing depth. This revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**. Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% across the samples **(Figure \ref{16scompare} B)**. Although the features in the viral model also were of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.

The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature also was driven by a few OVUs, suggesting a role for these viruses in tumorigenesis **(Figure \ref{predmodel} D)**. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and phage taxa who could not be confidently idetnfied beyond their broad phage identification (i.e. "unclassified"). Many of the important viruses, who were confirmed to not have genomic similarity to known bacterial genomes, were unidentifiable (denoted "unknown"). This is common in viromes across habitats; studies have reported as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

## Phage Influence Between CRC Stages
Because previous work has identified shifts in which bacteria were most important at different stages of colorectal cancer [@Zackular:2014fba; @Zackular:2016en; @Flynn:2016iu], we explored whether shifts in the relative influence of phages could be detected between healthy, adenomatous, and cancerous colons. We evaluated community shifts between the disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions. While bacterial OTU models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important between the healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.

@@

## Phage Dominance in CRC Virome
Differences in the colorectal cancer virome could have been driven by eukaryotic viruses or by bacteriophages. To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes. It is important to note that this could have been influenced by our methodological biases against enveloped viruses (more common of eukaryotic viruses than bacteriophage), due to chloroform and DNase treatment for purification.

@@

Given these findings, we hypothesized that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria. In other words, we hypothesized that the influential bacteriophages were community hubs (i.e. central members) within the bacteria and phage interactive network. We investigated the potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community **(Figure \ref{network} A)** [@Hannigan:2018gv]. The predicted interactions were then used to identify phage community hubs. We calculated the alpha centrality (i.e. measure of importance in the ecological network) of each phage OVU's connection to the rest of the network. The phages with high centrality values were defined as community hubs. Next, the centrality of each OVU was compared to its importance in the colorectal cancer classification model. Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer also were more likely to be community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities and had broad host ranges.

@@

Using our findings here, we have begun a path toward the delineation of the role the colonic virome plays in colorectal cancer in the form of a working hypothesis that will guide our future studies **(Figure \ref{modelsummary} A)**. We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states. By implementing a machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer. The colorectal cancer virome was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phages indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].

In addition to modeling the potential connections between virus communities, bacterial communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur. This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu], although it is important to note that there are also many other alternative hypotheses by which the system could be operating. We hypothesize that the process begins with broadly infectious phages in the colon lysing, and thereby disrupting, the existing bacterial communities. This shift opens novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize. Once the initial influential founder bacteria establish themselves in the epithelium, secondary opportunistic bacteria are able to adhere to the founders, colonize, and establish a biofilm. Phages may play a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process important for effective biofilm growth [@Rossmann:2015cj]. The oncogenic bacteria may then be able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response. As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria whose relative abundances were most influential. As the bacteria enter their oncogenic synergy with the epithelium, we conjecture that the phages continue mediating biofilm dispersal. This process would thereby support the colonized oncogenic bacteria by lysing competing cells and releasing nutrients to other bacteria in the form of cellular lysates. In addition to highlighting the likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities this model will guide future research investigations of the role the virome plays colorectal cancer.

At this time, our working hypothesis represents a conceptualization of areas for future work, which will be required to characterize the colorectal cancer microbiome at the functional, mechanistic level. There are many different ways in which this system can be operating, and our working hypothesis is one. For example, it is possible that the bacterial communities cause a change in the virome instead of the virome altering the bacterial communities. Reviewing every possible mechanism and characteristic of the system is beyond the scope of this manuscript, but the results we present provide a step toward the answer. Important future directions will include larger cohort human studies, further *in vitro* and *in vivo* mechanistic experimentation, and even attempts at community studies using absolute abundance values instead of relative abundance, which would allow for more accurate community dynamic modeling.

@@

https://github.com/SchlossLab/Hannigan_CRCVirome_mBio_2018

@@

## Ecological Network Analysis & Correlations
The ecological network of the bacterial and phage operational genomic units was constructed and analyzed as previously described [@Hannigan:2018gv]. Briefly, a random forest model was used to predict interactions between bacterial and phage genomic units, and those interactions were recorded in a graph database using *neo4j* graph databasing software (v2.3.1). The degree of phage centrality was quantified using the alpha centrality metric in the igraph CRAN package. A Spearman correlation was performed between model importance and phage centrality scores.

@@

# Acknowledgments
The authors thank the Schloss lab members for their underlying contributions, and the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. The authors also thank the participants in the study cohort.

@@

![*Basic Quality Control Metrics. A) VLP genomic DNA yield from all sequenced samples. Each bar represents a sample which is grouped and colored by its associated disease group. B) Sequence yield following quality control including quality score filtering and human decontamination. Dashed line shows rarefaction line in which all samples with lower sequence yield were excluded from downstream analysis.*\label{qualcontrol}]()

@@

![*Beta-diversity comparing disease states and the study negative controls which remained following rarefaction. Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test. Comparisons in which the intervals cross the zero mean difference line (dashed line) were not significantly different.*\label{betaogunegative}]()
